Avoiding tipping points in fisheries management: A Bayesian Non-Parametric approach

### 2014-06-23 manuscripts/manuscript.Rmd
small typo fixes

abstract: |
          Model uncertainty and limited data are fundamental
          challenges to robust management of human intervention in a 
          natural system.  These challenges
          are acutely highlighted by concerns that many ecological systems
          may contain tipping points, such as Allee population sizes.  
          Before a collapse, we do not know
          where the tipping points lie, if they exist at all.  Hence,
          we know neither a complete model of the system dynamics nor
          do we have access to data in some large region of state-space
          where such a tipping point might exist.  We illustrate
          how a Bayesian Non-Parametric (BNP) approach using a Gaussian Process (GP) prior
          provides a flexible representation of this inherent
          uncertainty. We embed GPs in a
          Stochastic Dynamic Programming (SDP) framework in order to make
          robust management predictions with both model uncertainty and
          limited data. We use simulations to evaluate this
          approach as compared with the standard approach of using model
          selection to choose from a set of candidate models.  We find that model selection erroneously
          favors models without tipping points -- leading to harvest policies
          that guarantee extinction. The GPDP 
          performs nearly as well as the true model and significantly
          outperforms standard appraoches. We illustrate this using examples
          of simulated single-species dynamics, where the standard model
          selection approach should be most effective, and find that it
          still fails to account for uncertainty appropriately and leads
          to population crashes, while management based on the GPDP 
          does not, since it does not underestimate the uncertainty
          outside of the observed data.

@@

<!-- Although much argument can be made
over the 'correct' objective function, e.g. how to account for the
social value of fish left in the sea vs. the commercial value of fish
harvested; see @Halpern2013 for further discussion of this issue.  We can
always compare model performance across multiple potential objectives.
The important point is that a decision-maker does not necessarily need
a model that provides the best mechanistic understanding or the best
long-term outcome, but rather the one that idelivers the best management
performance over the period of interest.
-->

@@

Dynamic optimization  requires characterizing the dynamics of a state variable (or variables), a control action, and a value function.
In this paper, we consider only a single state variable, so that the ideas can be described as simply as possible. That is, we focus on a class of one-dimensional models of population dynamics as a
'best-case' scenario for the parametric model performance relative to
the GP.   This is a best-case scenario for the parametric models because the underlying dynamics are always simuated
from one of the parametric models in the set, whereas in reality we never
have the "true" model among our candidates.  By choosing one-dimensional
models with few parameters, we limit the chance that poor performance
will be due to our inability to estimate parameters well, something
that becomes a more severe problem for higher-dimensional parametric
models. Further, the parametric models we consider are those most commonly
used in modeling stock-recruitment dynamics or to model sudden transitions
between alternative stable states.

@@

$$X_{t+1} = f(X_t) + \varepsilon$$

@@

We use inverse Gamma priors on both the lengthscale $\ell$ and $\sigma$, of the form

@@

Changing the intensity of the stochasticity or the distance between stable and unstable points
does not impact the performance of the GP relative to the optimal solution
given the true model and true parameters.  The parametric models are
more senstive to this difference.  Large noise relative to the distance between the stable and unstable
point increases the chance of a stochastic transition.  More precisely,
if we let $L = K - x_c$, then the probability of a transition in a given
window of time $T$ scales as

@@

(see @Gardiner2009 or @Mangel2006 for the derivation).
As a result, the impact of using a model which underestimates the risk of harvesting
past the critical point is much more severe, since this such a situation
occurs more often.  Conversely, with large enough distance between the
optimal escapement and unstable points relative to the noise, the chance
of a transition becomes vanishingly small and all models can be estimated 
near-optimally. Models that underestimate the cost incurred by stock
sizes fluctuating significantly below the optimal escapement level will
not perform poorly as long as those fluctuations are sufficently rare.

The GPDP is weakly influenced by increasing stochasticity or increasing
Allee effects over much of the range (Figure 7) Much larger stochasticity or higher Allee
effects make even the optimal solution without any model or parameter uncertainty
unable to harvest the population effectively (e.g. noise large enoug to violate the
self-sustaining criterion of @Reed1979). Only at the largest Allee effect
sizes are some replicates seen to deviate more from the optimal solution.

>>>>

abstract: |
  Model uncertainty and limited data are fundamental challenges to
  robust management of human intervention in a natural system.
  These challenges are acutely highlighted by concerns that
  many ecological systems may contain tipping points, such as
  Allee population sizes.  Before a collapse, we do not know
  where the tipping points lie, if they exist at all.  Hence,
  we know neither a complete model of the system dynamics nor
  do we have access to data in some large region of state-space
  where such a tipping point might exist.  We illustrate how
  a Bayesian Non-Parametric (BNP) approach using a Gaussian
  Process (GP) prior provides a flexible representation of this
  inherent uncertainty. We embed GPs in a Stochastic Dynamic
  Programming (SDP) framework in order to make robust management
  predictions with both model uncertainty and limited data. We
  use simulations to evaluate this approach as compared with the
  standard approach of using model selection to choose from a set
  of candidate models.  We find that model selection erroneously
  favors models without tipping points -- leading to harvest
  policies that guarantee extinction. The GPDP performs nearly as
  well as the true model and significantly outperforms standard
  approaches. We illustrate this using examples of simulated
  single-species dynamics, where the standard model selection
  approach should be most effective, and find that it still
  fails to account for uncertainty appropriately and leads to
  population crashes, while management based on the GPDP does
  not, since it does not underestimate the uncertainty outside
  of the observed data.

@@

<!-- Although much argument can be made
over the 'correct' objective function, e.g. how to account for the
social value of fish left in the sea vs. the commercial value of fish
harvested; see @Halpern2013 for further discussion of this issue.  We can
always compare model performance across multiple potential objectives.
The important point is that a decision-maker does not necessarily need
a model that provides the best mechanistic understanding or the best
long-term outcome, but rather the one that delivers the best management
performance over the period of interest.
-->

@@

Dynamic optimization  requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function.
In this paper, we consider only a single state variable, so that the
ideas can be described as simply as possible. That is, we focus on a
class of one-dimensional models of population dynamics as a 'best-case'
scenario for the parametric model performance relative to the GP.
This is a best-case scenario for the parametric models because the
underlying dynamics are always simulated from one of the parametric models
in the set, whereas in reality we never have the "true" model among our
candidates.  By choosing one-dimensional models with few parameters, we
limit the chance that poor performance will be due to our inability to
estimate parameters well, something that becomes a more severe problem
for higher-dimensional parametric models. Further, the parametric models
we consider are those most commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable states.

@@

$$X_{t+1} = f(X_t) + \varepsilon,$$

@@

We use inverse Gamma priors on both the length-scale $\ell$ and $\sigma$, of the form

@@

Changing the intensity of the stochasticity or the distance between stable and unstable points
does not impact the performance of the GP relative to the optimal solution
given the true model and true parameters.  The parametric models are
more sensitive to this difference.  Large noise relative to the distance between the stable and unstable
point increases the chance of a stochastic transition.  More precisely,
if we let $L = K - x_c$, then the probability of a transition in a given
window of time $T$ scales as

@@

(see @Gardiner2009 or @Mangel2006 for the derivation).
As a result, the impact of using a model which underestimates the risk of harvesting
past the critical point is much more severe, since this such a situation
occurs more often.  Conversely, with large enough distance between the
optimal escapement and unstable points relative to the noise, the chance
of a transition becomes vanishingly small and all models can be estimated 
near-optimally. Models that underestimate the cost incurred by stock
sizes fluctuating significantly below the optimal escapement level will
not perform poorly as long as those fluctuations are sufficiently rare.

The GPDP is weakly influenced by increasing stochasticity or increasing
Allee effects over much of the range (Figure 7) Much larger stochasticity or higher Allee
effects make even the optimal solution without any model or parameter uncertainty
unable to harvest the population effectively (e.g. noise large enough to violate the
self-sustaining criterion of @Reed1979). Only at the largest Allee effect
sizes are some replicates seen to deviate more from the optimal solution.
